An elaborate data set on human gait and the effect of mechanical perturbations

### 2014-12-17 paper.tex

\section*{Introduction}
%
The collection of dynamical data during human walking has a long history
beginning with the first motion pictures and now with modern marker based
motion capture techniques and high fidelity ground reaction load measurements.
Even though years of data on thousands of subjects now exist, this data is not
widely disseminated, well organized, nor available with few or no restrictions.
David Winter's published normative gait data, \cite{Winter1990}, is widely used
in biomechanical studies, yet it comes from relatively few subjects and only a
small number of gait cycles per subject. This small source has successfully
inspired many other studies, such as powered prosthetic control
design, \cite{Sup2008}, but success in other research fields using large sets
of data for discovery lead one to believe that more elaborate data sets may
benefit the field of human motion studies. To enable such work, biomechanical
data needs to be shared extensively, organized, and curated to enabled future
analysts.

There are some notable gait data sets and databases besides Winter's
authoritative set that are publicly available. The International Society of
Biomechanics has maintained a web page (http://isbweb.org/data) since
approximately 1995 that includes data sets for download and mostly unencumbered
use. For example, Vaughn, et. al's data, \cite{Vaughan1992}, with kinematics
and force plate measurement from several subjects is available on the site. At
another website, the CGA Normative Gait Database, \cite{Kirtley2014}, Chris
Kirtley shares normative gait data from several studies and these files have
influenced other studies, for example the average gait cycles from children
used in \cite{Bogert2003}.

Chester et. al, \cite{Chester2007}, report on a large gait database comparison
where one database contained kinematic data of 409 gait cycles of children from
1 to 7 years old but the data does not seem to be publicly available. This is
unfortunately typical. But Tirosh et. al, recognized the need for a
comprehensive data base for clinical gait data and created the
Gaitabase, \cite{Tirosh2010}. This database may contain a substantial amount of
data but it is encumbered by a very complicated and restrictive license and
sharing scheme. However, there are examples of data with less restrictions. The
University of Wisconsin at LaCrosse has an easily accessible normative gait
data set, \cite{Willson2014}, from 25 subjects with lower extremity marker data
from multiple gait cycles and force plate measurements from a single gait
cycle.

@@

Early on, we discovered that the factory setup of the R-Link treadmill had a
vibration mode as low as 5\si{\hertz} that is detectable in the force measurements,
likely due to the flexible undercarriage and pitch motion mechanism. Trials 6-8
are affected by this vibration mode. During trials 9-15 the treadmill was
stabilized with wooden blocks. During, the remaining trials the treadmill was
stabilized with metal supports. See the Data Limitations Section for more
details.

@@

The MATLAB script and Simulink model produce a comma-delimited text file of six
signals: time stamp, slow, normal, and fast walking perturbation signals, and
slow and fast running signals.~\footnote{The running signals were not used in
the experiments presented in this paper.} The measured speed of the treadmill
belts are compared to the control input signals in
Figure~\ref{fig:input_output} to show the effect of the treadmill and
controller dynamics. The system introduces a delay and seems to act as a low
pass filter. The standard deviations of the outputs do not significantly differ
from the desired values: \SIlist{0.05;0.12;0.2}{\meter\per\second} for the
three speeds, respectively.
%
\begin{figure}
  \centering
  \includegraphics{figures/input_vs_output.pdf}
  \caption{Treadmill belt speed input signals (purple) and recorded output
    speeds (blue) for average belt speeds of
    \SIlist{0.8;1.2;1.6}{\meter\per\second}, respectively.}
  \label{fig:input_output}
\end{figure}

@@

For more insight into the difference in the unperturbed and perturbed data,
Figure~\ref{fig:gait-cycle-stats-comparison} compares the distribution of a few
gait cycle statistics. One can see that the perturbed strides have a much
larger variation in frequency and length.
%
\begin{figure}
  \centering
  \includegraphics{figures/unperturbed-perturbed-boxplot-comparison.pdf}
  \cprotect\caption{Box plots of the average belt speed, stride frequency, and
    stride length which compare gait cycles for the unperturbed (purple) and
    perturbed (blue). Produced by \verb|src/unperturbed_perturbed_comparison.py|.}
  \label{fig:gait-cycle-stats-comparison}
\end{figure}

\subsection*{Data Limitations}
%
The data is provided in good faith with great attention to detail but as with
all data there are anomalies that may affect the use and interpretation of
results emanating from the data. The following list gives various notes and
warnings about the data that should be taken into account when making use of
it.
%
\begin{itemize}
  \item Be sure to read the notes in each meta data file for details about
    possible anomalies in that particular trial. Things such as marker dropout,
    ghost markers, and marker movement are the more prominent notes. Details
    about variations in the equipment on the day of the trial are also
    mentioned.
  \item The subject identification number 0 stands for "no subject" and was
    used whenever data was collected from the system with no subject on the
    treadmill, for example during the trials that were intended to be used for
    inertial compensation purposes. These trials play through the exact
    protocol as those with a human subject and the matching trials are
    indicated in the meta data. Matching unloaded trials were recorded on the
    same day as the loaded trials and is noted in the
    \verb|trial:files:compensation| section of the meta data file.
  \item Trials 1 and 2 were not recorded as part of this study. Those trial
    identification numbers were reserved for early data exploration from data
    collected in other studies.
  \item Trials 37, 38, and 39 do not exist. The numbers were accidentally
    skipped.
  \item Trials 9, 10, and 11 used a slightly different event definition where the
    calibration poses were not explicitly tagged by an event, yet the protocol
    was the identical to the following trials. The calibration pose will have
    to be determined manually.
  \item  Trials 6-15 have force measurements are affected by the treadmill
    vibration mode mentioned in the equipment section and the forces should be
    not be used. We include the trials because both the kinematic data is valid
    and trials 6-8 include lateral perturbations in addition to the
    longitudinal.
  \item During trials 9-15 we used wooden blocks to fix the treadmill to the
    concrete floor to eliminate the treadmill's low vibration mode
    (\~5\si{\hertz}). But these blocks seem to have corrupted the force plate
    measurements by imposing frictional stresses on the system. The force plate
    measurements should not be used from these trials, but the marker data is
    fine.
  \item Trials 6-8 use an early experimental protocol which divided the
    perturbation sections into three sections: longitudinal perturbations,
    lateral perturbations, and a combination of each. We then learned the
    treadmill had a low vibrational mode which significantly affects the force
    plate measurements, requiring us to eliminate the lateral perturbation
    motions. The force measurements during these trials are corrupted by this
    vibrational mode and should be used with caution or not at all.
  \item We did not record unloaded compensation trials for trials 9-15.
    Regardless, they would likely be useless due to the corruption from the
    wooden blocks.
  \item Trials 6-8 use a only the lower body marker set. The remaining trials
    are full body.
  \item The ankle joint torques computed from subject 9's data in trials 25-27
    are abnormal and should be used with caution or not at all. We were not
    able to locate the source of the error, but it is likely related to the
    force calibration.
\end{itemize}

>>>>

\section*{Introduction}
%
The collection of dynamical data during human walking has a long history
beginning with the first motion pictures and now with modern marker based
motion capture techniques and high fidelity ground reaction load measurements.
Even though years of data on thousands of subjects now exist, this data is not
widely disseminated, well organized, nor available with few or no restrictions.
David Winter's published normative gait data, \cite{Winter1990}, is widely used
in biomechanical studies, yet it comes from relatively few subjects and only a
small number of gait cycles per subject. This small source has successfully
inspired many other studies, such as powered prosthetic control
design, \cite{Sup2008}, but success in other research fields using large sets
of data for discovery lead one to believe that more elaborate data sets may
benefit the field of human motion studies. To enable such work, biomechanical
data needs to be shared extensively, organized, and curated to enable future
analysts.

There are some notable gait data sets and databases besides Winter's
authoritative set that are publicly available. The International Society of
Biomechanics has maintained a web page (http://isbweb.org/data) since
approximately 1995 that includes data sets for download and mostly unencumbered
use. For example, Vaughn, et. al's data, \cite{Vaughan1992}, with kinematics
and force plate measurement from several subjects is available on the site. At
another website, the CGA Normative Gait Database, \cite{Kirtley2014} shares
normative gait data from several studies and these files have influenced other
studies, for example the average gait cycles from children used in
\cite{Bogert2003}.

\cite{Chester2007} report on a large gait database comparison where one
database contained kinematic data of 409 gait cycles of children from 1 to 7
years old but the data does not seem to be publicly available. This is
unfortunately typical. But Tirosh et. al, recognized the need for a
comprehensive data base for clinical gait data and created the Gaitabase,
\cite{Tirosh2010}. This database may contain a substantial amount of data but
it is encumbered by a very complicated and restrictive license and sharing
scheme. However, there are examples of data with less restrictions. The
University of Wisconsin at LaCrosse has an easily accessible normative gait
data set, \cite{Willson2014}, from 25 subjects with lower extremity marker data
from multiple gait cycles and force plate measurements from a single gait
cycle.

@@

Early on, we discovered that the factory setup of the R-Link treadmill had a
vibration mode as low as 5\si{\hertz} that is detectable in the force measurements,
likely due to the flexible undercarriage and pitch motion mechanism. Trials 6-8
are affected by this vibration mode. During trials 9-15 the treadmill was
stabilized with wooden blocks. During the remaining trials the treadmill was
stabilized with metal supports. See the Data Limitations Section for more
details.

@@

The MATLAB script and Simulink model produce a comma-delimited text file of
five desired belt speed signals indexed by the time stamp. There are slow, normal,
and fast walking speeds and slow and fast running speeds.~\footnote{The running
signals were not used in the experiments presented in this paper.} The measured
speed of the treadmill belts are compared to these commanded treadmill control
input signals in Figure~\ref{fig:input_output} to show the effect of the
treadmill and controller dynamics. The system introduces a delay and seems to
act as a low pass filter. The standard deviations of the measured speeds do not
significantly differ from the desired values:
\SIlist{0.05;0.12;0.2}{\meter\per\second} for the three speeds, respectively.
%
\begin{figure}
  \centering
  \includegraphics{figures/input_vs_output.pdf}
  \caption{Treadmill belt speed input signals (purple) and recorded output
    speeds (blue) for average belt speeds of
    \SIlist{0.8;1.2;1.6}{\meter\per\second}, respectively.}
  \label{fig:input_output}
\end{figure}

@@

For more insight into the difference in the unperturbed and perturbed data,
Figure~\ref{fig:gait-cycle-stats-comparison} compares the distribution of a few
gait cycle statistics. One can see that the perturbed strides have a much
larger variation in frequency and length.
%
\begin{figure}
  \centering
  \includegraphics{figures/unperturbed-perturbed-boxplot-comparison.pdf}
  \cprotect\caption{Box plots of the average belt speed, stride frequency, and
    stride length which compare unperturbed (purple) and perturbed (blue) gait
    cycles. The median is given with the box bounding the first and third
    quartiles and the whiskers bound the range of the data. Produced by
    \verb|src/unperturbed_perturbed_comparison.py|.}
  \label{fig:gait-cycle-stats-comparison}
\end{figure}

\subsection*{Data Limitations}
%
The data is provided in good faith with great attention to detail but as with
all data there are anomalies that may affect the use and interpretation of
results emanating from the data. The following list gives various notes and
warnings about the data that should be taken into account when making use of
it.
%
\begin{itemize}
  \item Be sure to read the notes in each meta data file for details about
    possible anomalies in that particular trial. Things such as marker dropout,
    ghost markers, and marker movement are the more prominent notes. Details
    about variations in the equipment on the day of the trial are also
    mentioned.
  \item The subject identification number 0 stands for "no subject" and was
    used whenever data was collected from the system with no subject on the
    treadmill, for example during the trials that were intended to be used for
    inertial compensation purposes. These trials play through the exact
    protocol as those with a human subject and the matching trials are
    indicated in the meta data. Matching unloaded trials were recorded on the
    same day as the loaded trials and is noted in the
    \verb|trial:files:compensation| section of the meta data file.
  \item Trials 1 and 2 were not recorded as part of this study. Those trial
    identification numbers were reserved for early data exploration from data
    collected in other studies.
  \item Trials 37, 38, and 39 do not exist. The numbers were accidentally
    skipped.
  \item Trials 9, 10, and 11 used a slightly different event definition where the
    calibration poses were not explicitly tagged by an event, yet the protocol
    was the identical to the following trials. The calibration pose will have
    to be determined manually.
  \item  Trials 6-15 have force measurements are affected by the treadmill
    vibration mode mentioned in the equipment section and the forces should be
    not be used. We include the trials because both the kinematic data is valid
    and trials 6-8 include lateral perturbations in addition to the
    longitudinal.
  \item During trials 9-15 we used wooden blocks to fix the treadmill to the
    concrete floor to eliminate the treadmill's low vibration mode
    (approximately 5\si{\hertz}). But these blocks seem to have corrupted the
    force plate measurements by imposing frictional stresses on the system. The
    force plate measurements should not be used from these trials, but the
    marker data is fine.
  \item Trials 6-8 use an early experimental protocol which divided the
    perturbation sections into three sections: longitudinal perturbations,
    lateral perturbations, and a combination of each. We then learned the
    treadmill had a low vibrational mode which significantly affects the force
    plate measurements, requiring us to eliminate the lateral perturbation
    motions. The force measurements during these trials are corrupted by this
    vibrational mode and should be used with caution or not at all.
  \item We did not record unloaded compensation trials for trials 9-15.
    Regardless, they would likely be useless due to the corruption from the
    wooden blocks.
  \item Trials 6-8 use a only the lower body marker set. The remaining trials
    are full body.
  \item The ankle joint torques computed from subject 9's data in trials 25-27
    are abnormal and should be used with caution or not at all. We were not
    able to locate the source of the error, but it is likely related to the
    force calibration.
\end{itemize}
